---
title: 'Intro to Data Analysis in R: Session 2'
output: slidy_presentation
date: "October 20, 2017"
subtitle: UCI Data Science Initiative
smaller: yes
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Session 2 - Agenda

1. R Packages and Functions
2. Exploratory Data Analysis


## Functions and Packages:

1. R has many built-in functions
2. Each function has a name followed by (), e.g., ```mean()```
3. Arguments of a function are put within the parentheses
4. R packages are a way to maintain collections of R functions and data sets
5. Packages allow for easy, transparent and cross-platform extension of the R base system


## Functions and Packages:

Terminology:
  
  1. Package: an extension of the R base system with code, data and documentation in a standardized format
2. Library: a directory containing installed packages
3. Repository: a website providing packages for installation
4. Source: the original version of a package with human-readable text and code
5. Base packages: part of the R source tree, maintained by R Core

+ for more info on how R packages are developed, please read: "Creating R Packages: A Tutorial" (Friedrich Leisch)
+ http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf


## How to install a package in R:

There are two main ways to install a package in R:
  
  1. Installing from CRAN: install a package directly from the repository
+ Using R studio: tools/install packages
+ From R console: ```install.packages()```
2. Installing from Source: first download the add-on R package and then type the following in your console:
  + ```install.packages("path_to_file", repos = NULL, type = "source")``` 

+ Once you install a package, you need to load it into R using the function ```library()```


## Popular Packages in R:

1. To visualize data:
  + ggplot2: to create beautiful graphics
+ googleVis: to use Google Chart tools

2. To report results:
  + shiny: to create interactive web-based apps
+ knitr: to combine R codes and Latex/Markdown codes
+ slidify: to build HTML 5 slide shows

3. To write high-performance R code:
  + Rcpp: to write R functions that call C++ code
+ data.table: to organize datasets for fast operations
+ parallel: to use parallel processing in R



## Functions in R

+ Consider the function sample(). Run ?sample to read the help file or str(sample) to see its arguments. 
```{r echo=TRUE}
str(sample)
```

+ sample() has four arguments: 
  + x: vector of elements from which to choose
+ size: desired sample size
+ replace: sampling with/without replacement (logical)
+ prob: vector of probability weights

+ The help file will specify which arguments have default values (and what those values are)


## Calling a function in R

Function arguments can either be matched by position within the parentheses or by name

```{r echo=TRUE} 
sampSpace <- 1:6 
sample(sampSpace, 1) # arguments with default values can be omitted
sample(size = 1, x = sampSpace) # no need to remember the order 
sample(size = 1, sampSpace)
```

## Some Useful Functions:

+ Here we are going to talk about:
+ str(): a function to explain internal structure of an object
+ summary(): a function that summarizes variables in a data frame 
    + Note: this function is also used to summarize results of model fitting functions, which we will go over in the afternoon


## str():

+ str() is a compact way of understanding what an object is and what is in that object
```{r echo=TRUE}
str(str)
str(sample)
```

## str():

+ After loading a data frame, it is often useful to use ```str()``` in order to understand the structure of your data. 

```{r echo = TRUE, eval=TRUE}
prestige <- read.table(file = "prestige.csv", sep=",", header = TRUE)
str(prestige)
```
## summary():

+ ```summary()``` is another useful function for understanding your data

```{r echo = TRUE, eval=TRUE}
summary(prestige)
```





