---
title: "Intro to Data Analysis with R: Session 2"
subtitle: UCI Data Science Initiative
date: "October 20, 2017"
#author: "Chris Galbraith, Emily Smith"
output: slidy_presentation
smaller: yes
---

```{r, include=FALSE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Session 2 - Agenda
1. R Packages, Functions & Help

2. Control Structures

3. Exploratory Data Analysis

    + Basic Plotting
    + Histograms
    + Boxplots
    + Scatterplot Matrices
    

## Functions & Packages
1. R has many built-in functions
2. Each function has a name followed by (), e.g., ```mean()```
3. Arguments of a function are put within the parentheses
4. R packages are a way to maintain collections of R functions and data sets
5. Packages allow for easy, transparent and cross-platform extension of the R base system


## Functions and Packages
Terminology:
  
+ *Package*: an extension of the R base system with code, data and documentation in a standardized format
+ *Library*: a directory containing installed packages
+ *Repository*: a website providing packages for installation
+ *Source*: the original version of a package with human-readable text and code
+ *Base packages*: part of the R source tree, maintained by R Core

For more info on how R packages are developed, please read [Creating R Packages: A Tutorial](http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf) by Friedrich Leisch


## Installing a Package
There are two main ways to install a package in R:

1. Installing from CRAN: install a package directly from the repository
    + Using R studio: tools/install packages
    + From R console: ```install.packages()```
2. Installing from source: first download the add-on R package and then type the following in your console:
    + ```install.packages("path_to_file", repos = NULL, type = "source")``` 

Once you install a package, you need to load it into R using the function ```library()```


## Popular Packages
+ To visualize data:
    + ```ggplot2```: to create beautiful graphics
    + ```googleVis```: to use Google Chart tools
+ To report results:
    + ```shiny```: to create interactive web-based apps
    + ```knitr```: to combine R codes and Latex/Markdown codes
    + ```slidify```: to build HTML 5 slide shows
+ To write high-performance R code:
    + ```rcpp```: to write R functions that call C++ code
    + ```data.table```: to organize datasets for fast operations
    + ```parallel```: to use parallel processing in R


## Functions in R
+ Consider the function ```sample()```. 
+ Run ?sample to read the help file or str(sample) to see its arguments. 
```{r echo=TRUE}
str(sample)
```
+ ```sample()``` has four arguments: 
    + ```x```: vector of elements from which to choose
    + ```size```: desired sample size
    + ```replace```: sampling with/without replacement (logical)
    + ```prob```: vector of probability weights
+ The help file will specify which arguments have default values (and what those values are)


## Calling a Function in R
Function arguments can either be matched by position within the parentheses or by name

```{r echo=TRUE} 
sampSpace <- 1:6 
sample(sampSpace, 1) # arguments with default values can be omitted
sample(size = 1, x = sampSpace) # no need to remember the order 
sample(size = 1, sampSpace)
```


## Some Useful Functions
+ ```str()```: a function to explain internal structure of an object
+ ```summary()```: a function that summarizes variables in a data frame 
    + Note: this function is also used to summarize results of model fitting functions, which we will go over in the afternoon.


## ```str()```
Compact way of understanding what an object is and what is in that object
```{r echo=TRUE}
str(str)
str(sample)
```


## ```str()```
After loading a data frame, it is often useful to use ```str()``` in order to understand the structure of your data. 

```{r echo = TRUE, eval=TRUE}
prestige <- read.table(file = "./data/prestige_v2.csv", sep=",", header = TRUE)
str(prestige)
```


## ```summary()```
Another useful function for understanding your data
```{r echo = TRUE, eval=TRUE}
summary(prestige)
```


## Control Structures

+ A control structure is a block of programming that analyzes variables and chooses a direction in which to go based on given parameters
+ Control structures in R include:
    + for loops
    + if/else statements
    + while loops 
    + repeat
    + break 
    + next
    + return

## Exploratory Data Analysis

+ Before performing statistical analyses on your data, it is important to do exploratory data analysis (EDA) in order to better understand the variables and the relationships between them
+ Here we will cover some of the basic plotting functions in R

## Histogram

+ One of the most basic plots in R is a histogram
+ Let's plot a histogram of the response variable prestige
    + We can add arguments to the function to change the bar colors, title, and axis labels
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, col = "grey", main = "Histogram of Prestige Score", xlab = "Prestige Score")
```

## Histogram

+ We can use the function ```abline()``` to add a vertical line at the median of prestige
```{r echo = TRUE, eval=TRUE}
hist(prestige$prestige, col = "grey", main = "Histogram of Prestige Score", xlab = "Prestige Score")
abline(v = median(prestige$prestige), col = "blue")
```

## Scatterplot

+ We can use scatterplots to visualize the relationship between two variables, such as education and prestige
```{r echo = TRUE, eval=TRUE}
plot(prestige$education, prestige$prestige,
     main = "Prestige Score by Education",
     xlab = "Ave. Years of Education", ylab = "Prestige Score")
```

## Scatterplot

+ Use the functions ```lines()``` and ```lowess()``` to add a smoothed line through the plot
```{r echo = TRUE, eval=TRUE}
plot(prestige$education, prestige$prestige,
     main = "Prestige Score by Education",
     xlab = "Ave. Years of Education", ylab = "Prestige Score")
lines(lowess(prestige$education, prestige$prestige), col = "blue", lty = 2, lwd = 2)
```

## Scatterplot Matrix

+ The function ```scatterplotMatrix()``` (found in the car package) produces scatterplots between all variables in a dataframe
+ We can use direct ordering of the variables to control the order in which they are plotted

```{r echo = TRUE, eval=TRUE}
library(car)
scatterplotMatrix(prestige[,c("prestige","education","income","women")])
```

## Boxplot

+ The ```boxplot()``` function can be used to produce a boxplot of a single variable, or of a variable by categorical group
+ First let's make a boxplot of the response variable prestige

```{r echo = TRUE, eval=TRUE}
boxplot(prestige, data = prestige)
```

## Boxplot

+ Next we make a boxplot of prestige by the categorical variable "type"

```{r echo = TRUE, eval=TRUE}
boxplot(prestige ~ type, data = prestige, col = "grey",
        main = "Distribution of Prestige Score by Type of Occupation",
        xlab = "Occupation Types", ylab = "Prestige Score")
```