---
title: 'Intro to R Workshop: Session 4'
subtitle: UCI Data Science Initiative
date: "October 20, 2017"
#author: Chris Galbraith
output: slidy_presentation
smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Session 4 Agenda
+ Logistic Regression

    + Brief Intro to Generalized Linear Models (GLMs)
    + Fitting Models for Binary Response Data
    + Interpretation
    + Diagnostics
    + Prediction


## O-Ring Data
asdf


## Generalized Linear Models (GLMs)
asdf


## GLMs in R
asdf


##  Logistic Regression - O-Ring Data
+ `glm()` is used to fit generalized linear models (logistic regression in this example).
+ Logistic regression is used to model the probability of a binary outcome.
+ For instance, say we want to predict female labor force participation using the `Mroz` data.

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(car)
data(Mroz) # load Mroz data
str(Mroz)
```


## Logistic Regression - Data Description
+ lfp: female labor-force participation (factor with levels no/yes)
+ k5: number of children 5 years old or younger
+ k618: number of children 6 to 18 years old
+ age: in years
+ wc: wife's college attendance (factor with levels no/yes)
+ hc: husband's college attendance (factor with levels no/yes)
+ lwg: log expected wage rate
    + for women in the labor force, the actual wage rate
    + for women not in the labor force, an imputed value 
+ inc: family income exclusive of wife's income


## Logistic Regression - Exploratory Data Analysis (EDA)
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
summary(Mroz)
```


## Logistic Regression - EDA
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
barplot(table(Mroz$lfp), col = "blue", 
        main = "Count of Females by Labor Force Participation",
        xlab = "Labor Force Participation Status", ylab = "Number of Females")
```


## Logistic Regression - EDA
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
counts <- table(Mroz$lfp, Mroz$k5)
barplot(counts, col = c("blue","red"),
        main = "Labor Force Participation by # of Children < 5",
        xlab = "# of Children < 5", ylab = "Count",
        legend = rownames(counts))
```


## Logistic Regression - EDA
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
counts <- table(Mroz$lfp, Mroz$k5)
barplot(counts, col = c("blue","red"),
        main = "Labor Force Participation by # of Children < 5",
        xlab = "# of Children < 5", ylab = "Count",
        legend = rownames(counts), beside = T)
```


## Logistic Regression - EDA
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
plot(Mroz$age, Mroz$lwg, col = ifelse(Mroz$lfp == "yes", "red", "blue"),
     main = "Age vs. Log Wage by LFP", xlab = "Age (Years)", ylab = "Wage")
legend(x = 50, y = -1.5, legend = c("in LF", "Out of LF"),
       fill = c("red", "blue"), bty = "n")
```


## Logistic Regression - Model Fit
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
fitLogistic <- glm(lfp ~ k5 + age, family=binomial(logit), data=Mroz)
fitLogistic
```


## Logistic Regression - Model Fit
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
summary(fitLogistic)
```


## Logistic Regression - Model Fit
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
names(fitLogistic)
```


## Logistic Regression - Exponentiated CIs
+ 95% CI for exp(coefficients) (profile likelihood method)
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
exp(confint(fitLogistic, level=0.95))
```

+ 95% CI for exp(coefficients) (Wald confidence interval)
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
exp(confint.default(fitLogistic, level=0.95))
```


## Logistic Regression - Predictions
To get probabilities, be sure to use type = "response"
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
head(Mroz,1)
XB <- as.vector(head(predict(fitLogistic),1))
XB
p <- as.vector(head(predict(fitLogistic, type = "response"),1))
p
exp(XB)/(1 + exp(XB))  # inverse logit function
```


## Logistic Regression - Updates
+ Update model by adding 'inc' and 'lwg'
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
fitLogistic2 = update(fitLogistic, . ~ . + inc + lwg, data=Mroz)
```

+ After update
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
fitLogistic2
```


## Model Comparison
+ Use change of deviance of fitted model
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
anova(fitLogistic, fitLogistic2, test='LRT')
```


## End of Session 4
Make sure you submit course evaluation surveys!



