---
title: "Intro to Data Analysis with R: Exercise 2 Solutions"
subtitle: UCI Data Science Initiative
# author: "Chris Galbraith"
output: html_document
---

## Introduction
This set of exercises picks up where we left off with the Auto MPG data in Exercise 1. We now focus on exploratory data analysis.


### Part A 
**A.1** Open a new R script and save it in the directory you created in Part A.1 of Exercise 1. Then, load the Auto MPG data set with the additional variable ```diesel``` using the ```load()``` function and file path from the end of Exercise 1. 
```{r echo= TRUE}
load(here::here("data", "auto_mpg_v2.Rda"))
```

**A.2** Using the ```summary()``` function, look at descriptive statistics for the Auto MPG data. What do you notice?
```{r echo= TRUE}
summary(auto)
```

### Part B
**B.1** Plot a relative frequency histogram of the response variable, MPG. Color the boxes grey. Make sure to name the plot and axes.
```{r echo=TRUE}
hist(auto$mpg, freq=FALSE, col = "grey", main = "Histogram of MPG", xlab = "MPG")
```

**B.2** Add a density curve to the histogram you plotted in B.1 using the ```lines()``` and ```density()``` functions. Color it red using the ```col``` argument. HINT: You need to change the value of ```na.rm``` in the density function to avoid an error. You also need to make sure that the histogram is on the proper scale!
```{r echo=TRUE}
hist(auto$mpg, freq=FALSE, col = "grey", main = "Histogram of MPG", xlab = "MPG")
lines(density(auto$mpg, na.rm=TRUE), col="red")
```

**B.3** Add a vertical line to the plot from B.2 at the median of MPG. Make it dashed using the ```lty``` arguement, and make it red. HINT: You may need to use ```na.rm``` again.
```{r echo=TRUE}
hist(auto$mpg, freq=FALSE, col = "grey", main = "Histogram of MPG", xlab = "MPG")
lines(density(auto$mpg, na.rm=TRUE), col="red", main="Densities of Samp1 and Samp2", xlab="")
abline(v = median(auto$mpg, na.rm=TRUE), col = "red", lwd = 2, lty=2)
```


### Part C
Create a boxplot of MPG grouped by the number of cylinders. Color the boxes grey. Make sure to name the plot and axes.
```{r echo=TRUE}
boxplot(mpg ~ cyl, data = auto, col = "grey",
        main = "Distribution of MPG by Number of Cylinders",
        xlab = "Cylinders", ylab = "MPG")
```


### Part D
Create a scatterplot matrix of the Auto MPG data. Include the following variables: "mpg", "cyl", "disp", "hp", "weight", "acc", "model.yr". What relationships do you see? Is there any confounding? Which variables should be included in a regression? 
```{r echo=TRUE, warning=FALSE}
library(car)
scatterplotMatrix( auto[, c("mpg", "cyl", "disp", "hp", "weight", "acc", "model.yr")] )
```
