---
title: 'Intro to Data Analysis with R: Session 4'
subtitle: UCI Data Science Initiative
params: 
  reportdate: "x"  # to be replaced
date: "`r params$reportdate`"
#author: Chris Galbraith
output: slidy_presentation
smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Session 4 Agenda
+ [Motivating example](https://www.youtube.com/watch?v=j4JOjcDFtBE)
+ Brief Intro to Generalized Linear Models (GLMs)
+ Logistic Regression
    + Fitting Models for Binary Response Data
    + Interpretation
    + Diagnostics
    + Prediction


## O-Ring Data
+ On January 28, 1986, the USA Space Shuttle Challenger exploded 73 seconds into flight, killing all 7 crew members. The explosion was traced to failure of O-ring seals in the solid rocket booster at liftoff.

+ It was hypothesized that there is a inverse relationship between O-ring failure and temperature. 

+ The night before the flight, engineers had to make a decision whether or not to lauch Challenger the next day with a forecasted temperature of 31 degrees Fahrenheit.

+ The engineers had data available on 23 previous launches to aid in their decision with the following varaibles 

    + Number of O-rings at risk on a given flight 
    + Number experiencing thermal distress 
    + Launch temperature (degrees F) 
    + Leak-check pressure (psi) 
    + Temporal order of flight


## O-Ring Data, contd.
More information about the O-ring data is available [here](https://archive.ics.uci.edu/ml/datasets/Challenger+USA+Space+Shuttle+O-Ring).

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
oring <- read.table("https://archive.ics.uci.edu/ml/machine-learning-databases/space-shuttle/o-ring-erosion-only.data")
names(oring) <- c("n_risk", "n_fail", "temp", "psi", "order")
head(oring)
summary(oring)
```


## Visualizing the O-Ring Data
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
oring <- oring[order(oring$temp), ]  # sort by ascending temp
# some flights have the same number of failures & launch temp, so make point size vary 
pt.size <- rep(1, nrow(oring))
dups <- oring[duplicated(cbind(oring$temp, oring$n_fail)), c("n_fail", "temp")]
pt.size[as.numeric(rownames(dups))] <- 2
trips <- dups[duplicated(dups), ]
pt.size[as.numeric(rownames(trips))] <- 3
plot(oring$temp, oring$n_fail, pch = 20, cex = pt.size, 
     xlab = "Temperature (deg F)", ylab = "Number of Failures")  
```


## Naive Analysis of the O-Ring Data
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
## Fit a linear model
linearReg <- lm(n_fail ~ temp, data = oring)
summary(linearReg)
```


## Naive Analysis Results
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
## Predict the output for the Challenger launch day
launchDay <- data.frame(temp = 31)
predict(linearReg, launchDay, interval="predict")  # point estimate & 95% CI

## Plot the fit extrapolating to launch day
plot(oring$temp, oring$n_fail, pch = 20, cex = pt.size/2, 
     xlab = "Temperature (deg F)", ylab = "Num O-Ring Failures",
     xlim = c(30, 85), ylim = c(0, 5))
abline(reg = linearReg, col = "red", lwd = 2)
points(x=31, y=2.52, pch=4, col="black", lwd=2)
x <- seq(25,90,.1)
pred <- predict(linearReg, data.frame(temp=x), interval="predict")
lines(x, pred[,2], lty=2, col="red")
lines(x, pred[,3], lty=2, col="red")
```


## Generalized Linear Models (GLMs)
+ Linear regression clearly isn't the proper technique for this type of data, but GLMs can overcome these issues.

+ Flexible extension of linear regression that allows response variable to have non-Gaussian error distributions.

+ Instead of modeling the mean $\mu$ of the response variable directly as a linear combination of the predictors, we model a **link function** $g(\mu)$. $$g[E(Y)] = X \beta \Rightarrow E(Y) = g^{-1}(X \beta) $$

+ $Y$ must have a distribution from the *exponential family*, which includes the Normal, binomial, Poisson, gamma, and many others.

+ Choice of link function $g$ depends on the distributional form of $Y$. There can be multiple link functions for any singe distribution. 


## Logistic Regression
+ Specific GLM when dealing with binary response data $Y_i \in \{0,1\}$ 
$$Y_i \sim Bernoulli(p_i)$$
$$\Rightarrow f(Y|p) = \prod_{i=1}^{n} \left[ p_i^{y_i} (1-p_i)^{1-y_i} \right]$$
where $p_i \in [0,1]$ is the probability that $Y_i = 1$.

+ Link function must map values in $(-\infty, \infty)$ to the range $[0,1]$
    
    + **logit**, probit (normal CDF), etc

+ Logit link function: 
$g(p) = \log\left( \frac{p}{1-p} \right) = X\beta$

+ Inverse of logit function is the logistic function: 
$p = g^{-1}(X\beta) = \frac{e^{X\beta}}{e^{X\beta} + 1} = \frac{1}{1+e^{-X\beta}}$


## GLMs in R
+ ```glm()``` is used to fit generalized linear models.

+ The ```formula``` argument is equivalent to that of ```lm()``` used for linear regression

+ New argument: ```family```

    + Specifies the distribution of the response varaible and the link function
    + Logistic regression: ```family = binomial(link="logit")```
    
+ ```Summary()```, ```update()```, ```predict()```, variable selection methods, etc. all work the same


## Logistic Regression - Format Data
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
## Reformat the response variable
oring$fail <- as.numeric(oring$n_fail > 0)
head(oring)

plot(oring$temp, oring$fail, pch = 20, cex = pt.size,
     xlab = "Temperature (deg F)", ylab = "O-Ring Failure")
```


## Logistic Regression - Fit the Model
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
## Fit a GLM with logistic link
logisticReg <- glm(fail ~ temp, data = oring, family=binomial(link="logit"))
summary(logisticReg)
```


## Logistic Regression
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
names(logisticReg)
```


## Logistic Regression - Plot the Fit
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
plot(oring$temp, oring$fail, pch = 20, cex = pt.size,
     xlab = "Temperature (deg F)", ylab = "O-Ring Failure")
curve(predict(logisticReg, data.frame(temp=x), type="response"), add=TRUE, col="red", lwd=2)
```


## Logistic Regression - Prediction
To get probabilities, be sure to use ```type = "response"```.
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
launchDay
predict(logisticReg, launchDay, type="response")
```

```predict()``` simply computes the linear combination & evaulates the inverse link
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
XB <- as.numeric(coef(logisticReg)[1] + coef(logisticReg)[2]*31)
XB  # estimated log-odds of o-ring failure on launch day
exp(XB)/(1 + exp(XB))  # estimated probability of o-ring failure on launch day
```


## Logistic Regression - Exponentiated CIs
+ $\exp(\beta)$ is the multiplicative change in the odds of O-ring failure for a one degree increase in temperature.

+ 95% CI for $\exp(\beta)$ (profile likelihood method)
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
exp(confint(logisticReg, "temp", level=0.95))
```

+ 95% CI for $\exp(\beta)$ (Wald confidence interval)
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
exp(confint.default(logisticReg, "temp", level=0.95))
```


## Logistic Regression - Updates
+ Update model by adding leak check pressure
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
logisticReg2 = update(logisticReg, . ~ . + psi, data=oring)
summary(logisticReg2)
```


## Logistic Regression - Model Comparison
+ Likelihood Ratio Test instead of F-test like in Linear Regression
```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
anova(logisticReg, logisticReg2, test='LRT')
```


## End of Session 4
**Make sure you submit a [course evaluation survey](https://docs.google.com/forms/d/1hcL2hbUchrrSqjd-__2dZEotk7wK2_GrJj_DsK-rxds/edit)!**



